---
title: "Data Science I: Grocery Prices"
author: "Carlie Cann, Nolan Dermigny, Kevin Pickelman"
format: pdf
editor: visual
---

## Introduction:

This project uses data from the U.S. Bureau of Labor Statistics to explore grocery prices over the past 15 years. The goal is to understand how the cost of common food items has changed over time and better understand how inflation affects people's daily lives. Whether you're a student on a budget, a family planning meals, or a policymaker concerned with the cost of living, these changes impact us all.

## Loading the Packages

First, let's load some R packages that may be useful!

```{r}
#install.packages("tidyverse")
#install.packages("rvest")
#install.packages("dplyr")
library("tidyverse")
library("rvest")
library("dplyr")

```

## Loading & Tidying The Data

Now, we must load and tidy the data from the data set.

```{r}
#Carlie
page <- read_html("Bureau of Labor Statistics Data.html")

coffee <- page |>
  html_node("#table117") |>
  html_table(fill = TRUE)
 
coffee_clean <- coffee %>%
  filter_all(all_vars(. != "" & !is.na(.)))

ice_cream <- page |>
  html_node("#table66") |>
  html_table(fill = TRUE)
 
ice_cream_clean <- ice_cream %>%
  filter_all(all_vars(. != "" & !is.na(.)))

cookie <- page |>
  html_node("#table12") |>
  html_table(fill = TRUE)
 
cookie_clean <- cookie %>%
  filter_all(all_vars(. != "" & !is.na(.)))

peaches <- page |>
  html_node("#table74") |>
  html_table(fill = TRUE)
 
peaches_clean <- peaches %>%
  filter_all(all_vars(. != "" & !is.na(.)))

white_potatoes <- page |>
  html_node("#table80") |>
  html_table(fill = TRUE)
 
white_potatoes_clean <- white_potatoes %>%
  filter_all(all_vars(. != "" & !is.na(.)))

tomato <- page |>
  html_node("#table82") |>
  html_table(fill = TRUE)
 
tomato_clean <- tomato %>%
  filter_all(all_vars(. != "" & !is.na(.)))

```

```{r}
#Nolan
page <- read_html("Bureau of Labor Statistics Data.html")

beans <- page |>
  html_node("#table102") |>
  html_table(fill = TRUE)
 
beans_clean <- beans %>%
  filter_all(all_vars(. != "" & !is.na(.)))

sugar <- page |>
  html_node("#table104") |>
  html_table(fill = TRUE)
 
sugar_clean <- sugar %>%
  filter_all(all_vars(. != "" & !is.na(.)))

oranges <- page |>
  html_node("#table69") |>
  html_table(fill = TRUE)
 
oranges_clean <- oranges %>%
  filter_all(all_vars(. != "" & !is.na(.)))

wine <- page |>
  html_node("#table132") |>
  html_table(fill = TRUE)
 
wine_clean <- wine %>%
  filter_all(all_vars(. != "" & !is.na(.)))

utility_gas <- page |>
  html_node("#table137") |>
  html_table(fill = TRUE)
 
utility_gas_clean <- utility_gas %>%
  filter_all(all_vars(. != "" & !is.na(.)))

electric <- page |>
  html_node("#table135") |>
  html_table(fill = TRUE)
 
electric_clean <- electric %>%
  filter_all(all_vars(. != "" & !is.na(.)))


```

```{r}
#Kevin
page <- read_html("Bureau of Labor Statistics Data.html")

bread <- page |>
  html_node("#table5") |>
  html_table(fill = TRUE)
 
bread_clean <- bread %>%
  filter_all(all_vars(. != "" & !is.na(.)))

steaks <- page |>
  html_node("#table148") |>
  html_table(fill = TRUE)
 
steaks_clean <- steaks %>%
  filter_all(all_vars(. != "" & !is.na(.)))

eggs <- page |>
  html_node("#table55") |>
  html_table(fill = TRUE)
 
eggs_clean <- eggs %>%
  filter_all(all_vars(. != "" & !is.na(.)))

chicken_breast <- page |>
  html_node("#table153") |>
  html_table(fill = TRUE)
 
chicken_breast_clean <- chicken_breast %>%
  filter_all(all_vars(. != "" & !is.na(.)))

whole_milk <- page |>
  html_node("#table57") |>
  html_table(fill = TRUE)
 
whole_milk_clean <- whole_milk %>%
  filter_all(all_vars(. != "" & !is.na(.)))

gas <- page |>
  html_node("#table82") |>
  html_table(fill = TRUE)
 
gas_clean <- gas %>%
  filter_all(all_vars(. != "" & !is.na(.)))

```

## Analyzing The Data

Now, it is time for the fun part... analyzing!

```{r}
#Carlie

sweets <- bind_rows(
  ice_cream_clean %>% mutate(Item = "Ice Cream"),
  cookie_clean %>% mutate(Item = "Cookies")
) %>%
  rename(Year = 1, Price = 2) %>%
  mutate(Year = as.integer(Year), Price = as.numeric(Price))


ggplot(sweets, aes(x = as.numeric(Year), y = as.numeric(Price), color = Item)) +
  geom_line(size = 1.2) +
  labs(title = "Price Trends of Sweet Items (2009–2024)",
       x = "Year", y = "Price (USD)", color = "Item") +
  theme_minimal()


```

```{r}
#Nolan
```

```{r}
#Kevin
```
